\emeszida

### 2024-03-23 sigbovik/paper.tex
Add \emeszida command
Fill in section 3

@@

\title{\textcuneiform{𒅴𒋃𒀀}}

@@

\section{Technical Description}
(of the Language)

@@

- reproduce entire perceptron-full.eme in the text, without comments

>>>>

\usepackage{longtable}

@@

\newcommand{\emeszida}{\textcuneiform{𒅴𒋃𒀀}}

\title{\emeszida}

@@

Unfortunately, the hardware required to execute these programs (i.e. a living Babylonian mathematician) has not been adequately preserved, but we have managed to write a Python library which emulates it.\footnote{\url{github.com/MrLogarithm/emeszida}}

\section{Description of the Language}
Programs in \emeszida\ follow a tabular structure with three main sections: (i) a header, denoted by \texttt{dub} \textcuneiform{𒁾} (``tablet''), (ii) a sequence of instructions, and (iii) a colophon detailing the tablet's authorship.
Each instruction spans four columns, which we have taken to calling the \textit{arguments}, \textit{opcode}, \textit{destination}, and \textit{line number}. 
These columns are usually tab-separated, though in a few documents they are TAB-separated (using the cuneiform sign \texttt{TAB} \textcuneiform{𒋰}).
Instructions are grouped into blocks by means of horizontal lines.

\paragraph{Arguments}
An instruction's arguments may be numbers, register addresses, or a combination of the two. Numbers are encoded following standard Babylonian conventions \cite{TODO}, with \textcuneiform{𒋙} denoting the radix point which separates the integer part from a following fraction. There is an explicit representation for zero (\textcuneiform{𒑱}), making these tablets some of the earliest unambigiuous examples of the mathematical concept of zero.

A register address is denoted by the phrase niŋ$_2$.kas$_7$ $n$-kam \textcuneiform{𒃻𒋃}$n$\textcuneiform{𒄰} (``thing.account $n$-th'', ``the item in the $n$-th place''), where $n$ is any number. \textcuneiform{𒃻𒋃} expressions can be nested to perform a kind of pointer dereferencing: for example, if \textcuneiform{𒃻𒋃𒐕𒄰} (register 1) contains the value \textcuneiform{𒐗} (3), and \textcuneiform{𒃻𒋃𒐗𒄰} (register 3) contains the value \textcuneiform{𒐛} (7), then \textcuneiform{𒃻𒋃𒃻𒋃𒐕𒄰𒄰} will evaluate to 7 (the value in the register pointed to by \textcuneiform{𒃻𒋃𒐕𒄰}).

If an instruction has multiple arguments, these must be delimited by a wide space (distinct from the short space used to separate groups of digits within a number), or by one of the phrases a.na \textcuneiform{𒀀𒈾} ``to'' or i.na \textcuneiform{𒄿𒈾} ``from''. By convention, the choice of delimiter depends on the instruction's opcode (see below), with multiplication operations preferring space delimiters, addition preferring \textcuneiform{𒀀𒈾}, and subtraction preferring \textcuneiform{𒄿𒈾}. 
There is no mechanism to enforce these conventions, but we recommend following them because \emeszida\ is hard enough to read at the best of times.

\paragraph{Destination}
In most cases, the destination column of an instruction will be a register address where the output is to be stored, e.g. \textcuneiform{𒃻𒋃𒐕𒄰}. 
Some control-flow instructions (see below) instead expect the destination to be a line number. 
\textcuneiform{𒋤} SUD (``distant, remote'') can be used as a null destination, for statements which produce no output.

\paragraph{Opcodes}
Each instruction has a single opcode belonging to the following vocabulary:
\begin{itemize}
    \item \textcuneiform{𒈭𒄩} DAḪ.ḪA, ``add''
    \item \textcuneiform{𒁀𒍣} BA.ZI, ``tear out''
    \item \textcuneiform{𒀀𒁺} A.RA$_2$, ``multiply''
    \item \textcuneiform{𒅆} IGI, ``reciprocal''
    \item \textcuneiform{𒈨} ME, ``to be''
    \item \textcuneiform{𒅆𒂃} \textit{ta-mar}, ``you will see''
    \item \textcuneiform{𒇔𒈾} NIGIN.NA, ``start again''
    \item \textcuneiform{𒋗𒌉𒁉𒋛𒀀} \textit{tukum.bi dirig}, ``if it exceeds''
    \item \textcuneiform{𒋗𒌉𒁉𒋝}  \textit{tukum.bi sig}, ``it it is weak''
\end{itemize}

The first three of these are binary operators for addition, subtraction, and multiplication respectively. 
The subtraction operator deserves special attention for its use in constructions of the following shape, which appear hundreds of times throughout the Babylonian programming corpus:

\begin{center}
    \textcuneiform{𒑱𒋙𒐖𒌍𒄿𒈾𒑱}\qquad\textcuneiform{𒁀𒍣}\qquad\textcuneiform{𒃻𒋃𒐕𒄰}\qquad\textcuneiform{𒑱}
\end{center}

This instruction subtracts \textcuneiform{𒑱𒋙𒐖𒌍} from zero and stores the result in \textcuneiform{𒃻𒋃𒐕𒄰}. 
This effectively negates the first argument, and seems to have been the primary way by which Babylonian cuneigrammers represented negative numbers, as their primitive and archaic notation otherwise lacked a means to encode such values.
This curious practice gives definitive proof that the invention of negative numbers occurred centuries earlier than previously believed.

The language does not appear to have any kind of binary division operator. Rather, a unary \textcuneiform{𒅆} operator was used to find the reciprocal of the denominator, which was then multiplied by the numerator using the binary \textcuneiform{𒀀𒁺} operator.

\textcuneiform{𒈨} is a unary assignment operator which stores a value in a destination register. \textcuneiform{𒅆𒂃} functions as a unary print operator.

\emeszida\ includes three types of control-flow instructions. \textcuneiform{𒇔𒈾} functions like GOTO, and jumps the program counter to the specified line number. \textcuneiform{𒋗𒌉𒁉𒋝} functions like the x86 \texttt{jz} instruction, and jumps to the specified line number if and only if its argument is zero. \textcuneiform{𒋗𒌉𒁉𒋛𒀀} is similar, and jumps if the argument is greater than zero.

\paragraph{Line Numbers}
Every line of an \emeszida\ program ends in a mandatory line number. 
However, these numbers are not generally sequential, and need not even be distinct. 
For example, most lines of the perceptron tablet are labeled with the number \textcuneiform{𒑱} (zero); only lines that are the destination of some control-flow instruction receive non-zero identifiers.

\subsection{Fractional Indexing}
Both line numbers and register addresses in \emeszida\ can have fractional parts.
The original scribes seem to have exploited this fact to establish non-overlapping ``namespaces'' for the different parts of their code. 
For example, in the perceptron tablet, all of the model parameters are stored in adresses with integer part 1; the program inputs all have integer part 2; the matrix multiplication subroutine uses addresses with integer part 3; and so on.
The fractional parts of register addresses also appear to follow some standard conventions, with the X;0 register typically storing a subroutine's return address, while X;1 onward were used for its arguments.

The perceptron tablet also uses fractional register addresses to perform a kind of multi-dimensional array indexing.
As an example, the first layer of the perceptron has a $50\times 2$ weight matrix, and this is stored in registers 1;0,0,0 through 1;0,49,1. The integer part of these addresses denotes the ``data'' portion of memory; the first digit after the radix point identifies this as the 0th model parameter; and the second and third digits can be treated as a pair of indices ranging from 0--49 and 0--1 respectively.
To access a specific element in this matrix, the scribes use repeated division by 60 to implement a kind of ``bit-shift'' instruction, in order to shift integer indices into the correct positions after the radix point. 
By adding the bit-shifted element indices (e.g. 0;0,4,7 for the element in row 5, column 8) to a pointer to the top corner of the matrix (1;0,0,0) they obtain the address of the desired element (1;0,4,7).

Notably, this practice limits the size of their model parameters to at most $60\times 60$, as for larger values the addresses would carry over to higher digits and thus begin to overwrite one another. 
This limitation may explain why AI never made waves in Babylonian society, as their models were all too small to be truly revolutionary.

\section{Description of the Texts}
- add graph plotting perceptron output vs true trigonometric ratios from Plimpton

@@

- reproduce entire perceptron-full.eme in the text, without comments:

\onecolumn
\tiny
\noindent
\input{perceptron.tex}
\normalsize
